Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-13 main/MSstatsPTM_Main.tex
Finalized main and supp text

\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}
\label{sec:comp_sim_procedure2}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}

>>>>

\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
\label{sec:comp_sim_procedure2}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low features}

### 2022-02-13 supplementary/ptm_sm.tex
Finalized main and supp text

\usepackage{geometry}
%\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{rotating}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,backgrounds,calc,positioning,fit,petri,plotmarks}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{authblk}
\usepackage{xr}
\externaldocument[main-]{../main/MSstatsPTM_Main}

@@

\subsection{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM. The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. \sfigref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein. Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization. To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

@@

The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis. To make the calculation the standard deviation and degrees of freedom are needed. The estimation of these quantities is dependent on the experimental design. In \sfigref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.

@@

%%%% Summary of controlled mixtures
\begin{table}
\begin{tiny}
\begin{tabular}{|c|c|ccccc|c|}
\hline
%%%% colnames of table
&  & \multicolumn{5}{c|}{Experimental design} & Data availability \\
& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& No. of Tech.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
&  & & Conditions & Replicates & Replicates &  & or R package) \\
\hline
\hline
% &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{3}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{3}{*}{\rotatebox[origin=c]{90}{ Ground}} \multirow{3}{*}{\rotatebox[origin=c]{90}{ Truth}} & DDA:ControlledMix \cite{Chiva2014} & Group & 5 & 1 & 3 & Suppl. Sec. & \\
 &&&&&&& \\[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& DDA:Choi2017 \cite{Choi:2017jan}  & Group & 4 & 1 & 3 &  & \\
 &&&&&&& \\[-0.05in]
%\hline 
% &&&&&&& \\[-0.05in]
%%% DDA Cox2014
& DDA:Cox2014 \cite{Cox:2014fi} & Group & 2 & 1 & 4 &  & \\
 &&&&&&& \\[-0.05in]
\hline
\multicolumn{8}{c}{ } \\ [0.02in]
\hline 
 &&&&&&& \\[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{investigation}} & DDA:Meierhofer2016 \cite{Meierhofer2016} & Paired & 2 & 6 & 2 &  &  \\
&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& SRM:Picotti2009  \cite{Picotti09} & Time course & 10 & 3 & 1 &  & \\
&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& SRM:Surinova2015:Training \cite{Surinova:2015jn} & Group & 2 & 100 & 1 & &\ [0.02in]
\end{tabular}\\
\end{tiny}
\caption{
}
\label{tab:dataDescriptionSpikein}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
\label{dataset1}
In the first simulation an experiment with many features per PTM and unmodified protein was created. Additionally this simulation contained no missing data. These attributes are not representative of a real experiment, but provide a baseline for model performance.

@@

The results are summarized from \sfigref{fig:sim1_fpr} to \sfigref{fig:sim1_recall}. These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy. Additionally, the results are compared to simulation 2 described in the next section.

@@

\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}

@@

The results are summarized from \sfigref{fig:sim2_recall} to \sfigref{fig:fc_boxplot}.

@@

Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test. All methods are analyzed after adjusting for changes in overall protein level. The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities. The results are summarized from \sfigref{fig:spike_volcano_msstats} to \sfigref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.

@@

The results of the proposed method to this experiment can be seen in \sfigref{fig:ipah_figures}.

@@

The results of the proposed method to this experiment can be seen in  \sfigref{fig:No_Diff_Shigella_PTM} and \sfigref{fig:Diff_Shigella_PTM}.

@@

An example profile plot for this experiment can be seen in  \sfigref{fig:USP30_profile_plot}.

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{sim_new/same_var_power}
	\caption{The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides. Predictably when the replicates are high for both modified and unmodified peptides the power is much higher. Conversely at low replicates for each the power is much lower. The interesting part of this chart is when the replicates are different between runs. With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
	\caption{In this chart the variance for the PTM is higher than the unmodified protein. The PTM variance is .2, while the unmodified protein variance is .1. With equal replicates the results are the same as above, more replicates equals more power. When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
 \end{subfigure}
 \caption{Power analysis of experiments with differing variances.}
\label{fig:power_sd_combo}
\end{figure}

>>>>

\usepackage{geometry}
%\usepackage{amsmath}
%\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{rotating}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,backgrounds,calc,positioning,fit,petri,plotmarks}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{url}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{authblk}
\usepackage{xr}
\usepackage{hyperref}
\externaldocument[main-]{../main/MSstatsPTM_Main}

@@

\subsection{Data structure of PTM quantification experiments.}
A set of fully-cleaved and/or partially-cleaved peptides containing a same PTM (e.g., ubiquitination/phosphorylation) at one site are considered together. There are $I$ conditions and $J$ mass spectrometry runs (technical replicates) per condition in the experiment. The PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. To account for the underlying protein abundance, features corresponding to the unmodified peptides from the same protein are considered together, except those unmodified peptides containing a modified site to avoid the confounding effect due to the PTM. The log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. Figure \ref{fig:dtable} shows an example data representation of modified peptide ions at one site and unmodified peptide ions of the same protein. Unmodified peptides from the same protein provide additional evidence on the underlying protein abundance, which needs to be integrated for PTM characterization. To address the goals of PTM characterization, statistical analysis needs to summarize values in this table using appropriate statistical models, translate the goal into a model-based quantity of interest, and draw inference (i.e., characterize the uncertainty) about the quantity.

@@

The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis. To make the calculation the standard deviation and degrees of freedom are needed. The estimation of these quantities is dependent on the experimental design. In Figure \ref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.

@@

Table \ref{tab:dataDescription} summarizes the datasets used in this manuscript and their experimental designs. There were 6 total experiments, 3 for which ground truth was known and 3 biological investigations. For the ground truth experiments, 2 were generated via computer simulation with differing parameters and the other was an experiment with heavy labeled spike-in peptides. The biological investigations represent a variety of experimental designs and acquisition methods. Two were acquired via Tandem Mass Tag (TMT) labeling, while the other was a label free experiment. Of the two TMT experiments, one was analyzed as a time series and the other a group comparison.

Details of raw data, R scripts with MSstatsPTM analyses, and result of statistical analyses for the datasets are available in either MassIVE.quant or Github (with direct links available in the table).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\begin{tiny}
\begin{tabular}{|c|c|cccc|c|}
\hline
%%%% colnames of table
&  & \multicolumn{4}{c|}{Experimental design} & Data availability \\
& Dataset & \multirow{2}{*}{Comparison} & No. of & No. of Bio.& \multirow{2}{*}{Details in}   & (MassIVE.quant  \\
&  & & Conditions & Replicates &  & or R package) \\
\hline
\hline
 &&&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset1} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group & 2/3/4 & 2/3/5/10 & Suppl. Sec. \ref{sec:dataset2} & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark & Group & 4 & 2 & Suppl. Sec. \ref{sec:benchmark} & \\
 &&&&&& \\%[-0.05in]
\hline
\multicolumn{7}{c}{ } \\ [0.02in]
\hline 
 &&&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - TMT & Group & 6 & 2 & Suppl. Sec. \ref{sec:ipah} &  \\
&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - TMT & Time course & 6 & 2 & Suppl. Sec. \ref{sec:shigella} & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group & 4 & 2 & Suppl. Sec. \ref{sec:usp30} &\\ [0.02in]
 &&&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset used in this manuscript.
`Comparison'' is the type of experimental design (group comparison, or time course).
``No. of Conditions'' states the number of conditions in the dataset.
``No. of Bio. Replicates'' states the number of biological replicates per condition.
Details regarding the experimental design for each dataset are available in the section listed in the ``Details in'' column.
``Data availability'' column shows the MassIVE.quant ID or location on github. 
}
\label{tab:dataDescription}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Dataset 1 : Computer simulation 1 - Label-free}
\label{sec:dataset1}
In the first simulation an experiment with many features per PTM and unmodified protein was created. Additionally this simulation contained no missing data. These attributes are not representative of a real experiment, but provide a baseline for model performance.

@@

The results are summarized from Figure \ref{fig:sim1_fpr} to Figure \ref{fig:sim1_recall}. These results include the false discovery rate with and without adjusting for changes in unmodified protein abundance, false positive rate, and overall accuracy. Additionally, the results are compared to simulation 2 described in the next section.

@@

\subsubsection{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
\label{sec:dataset2}

@@

The results are summarized from Figure \ref{fig:sim2_recall} to Figure \ref{fig:fc_boxplot}.

@@

Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test. All methods are analyzed after adjusting for changes in overall protein level. The proposed method summarizes feature intensities up to the run level using Tukey's Median Polish, while the other methods use the log of summed feature intensities. The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_fc}, including volcano plots, model summary statistics, and fold change analysis.

@@

The results of the proposed method to this experiment can be seen in Figure \ref{fig:ipah_figures}.

@@

The results of the proposed method to this experiment can be seen in Figure \ref{fig:No_Diff_Shigella_PTM} and Figure \ref{fig:Diff_Shigella_PTM}.

@@

An example profile plot for this experiment can be seen in  Figure \ref{fig:USP30_profile_plot}.

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{sim_new/same_var_power}
	\caption{The power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides. Predictably when the replicates are high for both modified and unmodified peptides the power is much higher. Conversely at low replicates for each the power is much lower. The interesting part of this chart is when the replicates are different between runs. With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
 \end{subfigure}\vspace{-1mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.75\textwidth]{sim_new/high_ptm_var_power}
	\caption{In this chart the variance for the PTM is higher than the unmodified protein. The PTM variance is .2, while the unmodified protein variance is .1. With equal replicates the results are the same as above, more replicates equals more power. When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
 \end{subfigure}\vspace{-1mm}
 \caption{Power analysis of experiments with differing variances.}
\label{fig:power_sd_combo}
\end{figure}
